---
title: "Analysis of gene-protein separation in cancer cells"
author: "Mahmoud Ahmed"
date: "11/30/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.width = 5, fig.height = 5, fig.align = 'center')
```

```{r load_libraries}
# load required libraries
library(tidyverse)
library(reshape2)
library(clusterProfiler)
library(cowplot)
library(jtools)
library(lme4)
library(nlme)
```

```{r ggplot_theme}
# define ggplot theme
my_theme <- theme(panel.grid = element_blank(),
                  panel.border = element_rect(fill = NA, color = 'black', size = 1),
                  panel.background = element_blank(),
                  axis.title.x.top = element_blank(),
                  axis.text.x.top = element_blank(),
                  axis.title.y.right = element_blank(),
                  axis.text.y.right = element_blank(),
                  legend.background = element_blank(),
                  axis.ticks.length=unit(2, "mm"))
```

# Datasets

The datasets used in this manuscript are the following:

- `proteins`: protein level data of different cancer cell lines
- `genes`: gene expression data of the same cell lines
- `growth`: division rate and growth inhibition data of a subset of the cell lines under drug perturbations
- `hallmarks`: lists of genes involved in the different cancer hall marks

The code in `code/prepare_data` download and tidy the data sets

```{r prepare_data, eval=FALSE}
# download and prepare the datasets
# source('code/prepare_data.R')
```

```{r load_data}
# load data
proteins <- read_rds('data/breast_cells_proteins.rds')
genes <- read_rds('data/breast_cells_genes.rds')
growth <- read_rds('data/breast_cells_growth.rds')
hallmarks <- read_rds('data/hallmarks_sets.rds')

# extract cells and symbols in common
common_cells <- intersect(colnames(genes), colnames(proteins))
common_symbols <- intersect(rownames(genes), rownames(proteins))

# subset genes and proteins
genes <- genes[common_symbols, common_cells]
proteins <- proteins[common_symbols, common_cells]

# test all have the same dims
dim(genes); dim(proteins)

# test all have the same names
all(colnames(genes) == colnames(genes))
all(rownames(genes) == rownames(proteins))
```

The following show the names and numbers of gene products, cell lines and drugs in the datasets.

```{r show_numbers}
# the number of cell lines
ncol(genes); ncol(proteins)
colnames(genes)

# the number of gene products
nrow(genes); nrow(proteins)

# the number of drugs
length(unique(growth$Small.Molecule.Name))
```

# Analysis

This analysis aim to

- Explore the correlation between gene and protein levels in cancer cell lines
- Show that these correlations systematically differ for cancer hallmark gene products
- Establish a correlation between gene-protein separation of the hallmarks and division rates under normal conditions and under drug perturbation
- Show how the range of enrichment associate with the same growth variables

## The correlation between gene and protein levels

```{r gene_protein_correaltion}
# calculate the correlation between genes and proteins in all cells
gene_protein_corr <- cor(genes, proteins, use = 'complete')

# make a tidy data.frame, and group by same or other cells
gene_protein_corr <- melt(gene_protein_corr) %>%
  mutate(group = ifelse(Var1 == Var2, 'Same cell', 'Other cells')) %>%
  as_tibble()
```

```{r gene_protein_corr_plot}
# plot the correlations within cell type and in between
corr_groups <- split(gene_protein_corr$value, gene_protein_corr$group)
corr_groups_ttest <- t.test(corr_groups$`Same cell`,
                            corr_groups$`Other cells`,
                            alternative = 'greater')

gene_protein_corr_plot <- gene_protein_corr %>%
  mutate(group = str_split(group, ' ', simplify = TRUE)[,1]) %>%
  ggplot(aes(x = value, fill = group)) +
  geom_bar(stat="bin", aes(y=..density..), alpha = .5) +
  scale_x_continuous(sec.axis = dup_axis(), limits = c(.09, .41)) +
  scale_y_continuous(sec.axis = dup_axis()) +
  geom_vline(xintercept = corr_groups_ttest$estimate,
             color = c('blue', 'red'),
             lty = 2) +
  geom_segment(aes(x = corr_groups_ttest$estimate[1],
                   xend = corr_groups_ttest$estimate[2],
                   y = 13, yend = 13),
               arrow = arrow(length = unit(2, 'mm'))) +
  geom_segment(aes(xend = corr_groups_ttest$estimate[1],
                   x = corr_groups_ttest$estimate[2],
                   y = 13, yend = 13),
               arrow = arrow(length = unit(2, 'mm'))) +
  annotate('text', x = .24, y = 16, size = 3, hjust = 0,
           label = paste('t =', round(corr_groups_ttest$statistic, 1))) +
  annotate('text', x = .24, y = 14.5, size = 3, hjust = 0,
           label = paste('P < 0.0001')) +
  labs(x = 'Gene-protein Correlation',
       y = 'Percent of cell lines',
       fill = '') +
  theme(legend.position = c(.2, .9),
        legend.background = element_blank()) +
  my_theme

gene_protein_corr_plot


plot_grid(gene_protein_corr_plot,
          hallmark_correlation_plot,
          nrow = 1,
          scale = .9,
          labels = 'AUTO',
          label_fontface = 'plain') %>%
  ggsave(plot = .,
         filename = 'output/manuscript/figures/gene_protein_correlation.png',
         width = 6, height = 3)
```

The figure shows that the gene protein correaltion is higher within cells, compared to between cells.

```{r average_correlation}
# average gene protein correlation
gene_protein_corr %>%
  group_by(group) %>%
  summarise(mean = mean(value),
            min = min(value),
            max = max(value))
```

```{r test_correlations}
# test the correlation within cell type and in between
corr_test <- ks.test(corr_groups$`Other cells`,
                     corr_groups$`Same cell`)
corr_test
```

The maximum distance between the two curves is `r round(corr_test$statistic, 2)`
The p-value of the test is `r corr_test$p.value`

```{r gene_protein_test_plot}
# generate ecdf curves plots
## calculate the ecdf for each group
ccdf1 <- ecdf(corr_groups$`Same cell`) 
ccdf2 <- ecdf(corr_groups$`Other cells`)

## get the minimum and maximum of the cumulative percentages
minMax <- seq(min(corr_groups$`Same cell`, corr_groups$`Other cells`),
              max(corr_groups$`Same cell`, corr_groups$`Other cells`),
              length.out=length(corr_groups$`Same cell`)) 

## extract the position of the maximum difference
xx0 <- minMax[which( abs(ccdf1(minMax) - ccdf2(minMax)) == max(abs(ccdf1(minMax) - ccdf2(minMax))) )] 
yy0 <- ccdf1(xx0) 
yy1 <- ccdf2(xx0) 

## generate a title
tt1 <- paste0(
  'D^+ = ',
  round(corr_test$statistic, 2),
  '; P ',
  ifelse(corr_test$p.value < .0001,
         '< 0.0001',
         paste0('= ', round(corr_test$p.value, 3)))
)

# make a plot
gene_protein_test_plot <- gene_protein_corr %>%
  ggplot() +
  stat_ecdf(aes(x = value, color = group)) +
  geom_segment(aes(x = xx0[1], y = yy0[1], xend = xx0[1], yend = yy1[1]),
               linetype = "dashed" ) +
  geom_point(aes(x = xx0[1] , y= yy0[1]), size=3) +
  geom_point(aes(x = xx0[1] , y= yy1[1]), size=3) +
  ggtitle(tt1) + 
  labs(x = 'Gene-protein Correlation',
       y = 'ECDF',
       color = '') +
  theme(legend.position = c(.25,.85),
        legend.background = element_blank())

gene_protein_test_plot
```

The figure visualizes the results of th KS test.

## Gene-protein correlations in cancer hallmakr gene sets

Here, we calculate similar correlation coefficients (Pearsons) for the gene sets involved in cancer hallmarks in each cell line.

```{r pearson_by_hallmark}
# calculate the pearsons correlation for the gene products in each hallmark
gene_protein_hallmark <- map_df(
  hallmarks,
  function(x) {
    as_tibble(
      melt(
        cor(
          genes[intersect(rownames(genes), x),],
          proteins[intersect(rownames(proteins), x),],
          use = 'complete'
        )
      )
    )
  }, .id = 'hallmark') %>%
  filter(Var1 == Var2)

head(gene_protein_hallmark)
```

Also, we calculate the Spearman rank correlations.

```{r spearman_by_hallmark}
# calculate the spearman correlation for the gene products in each hallmark
gene_protein_hallmark_spearman <- map_df(
  hallmarks,
  function(x) {
    as_tibble(
      melt(
        cor(
          genes[intersect(rownames(genes), x),],
          proteins[intersect(rownames(proteins), x),],
          use = 'complete',
          method = 'spearman'
        )
      )
    )
  }, .id = 'hallmark') %>%
  filter(Var1 == Var2)

head(gene_protein_hallmark_spearman)
```

Then we generate a random gene set of 500 gene products and calculate similar correlation coefficients.

```{r random_set}
# generate a random set of genes
set.seed(123)
random_set <- sample(rownames(genes), 500)
```

```{r pearson_in_random}
# calculate the correlation for the random set
gene_protein_random <- cor(
  genes[intersect(rownames(genes), random_set),],
  proteins[intersect(rownames(proteins), random_set),],
  use = 'complete'
) %>%
  melt() %>%
  as_tibble() %>%
  filter(Var1 == Var2) %>%
  mutate(hallmark = 'random')

head(gene_protein_random)
```

Then, we used a linear model to compare the average gene-protein correlation in the hallmarks with the random gene set.

```{r test_random_corr}
# test the correlation in the random set
## split by hallmark
hallmark_groups <- split(gene_protein_hallmark$value,
                         gene_protein_hallmark$hallmark)

## add random set
hallmark_groups$random <- gene_protein_random$value

## make a data.frame
hallmark_groups_df <- melt(hallmark_groups) %>%
  mutate(L1 = relevel(factor(L1), ref = 'random'))

## apply lm
cor_lm <- lm(value ~ L1, data = hallmark_groups_df)
summary(cor_lm)
```

Next, we applied analysis of variance to the resulting linear model

```{r corr_annova}
# apply annova
corr_anova <- anova(cor_lm)
corr_anova
```

The F-statistics is `r corr_anova$'F value'[1]` and the p-value is `r corr_anova$'Pr(>F)'[2]`

```{r corr_lm_plot}
# make a plot for the linear effects on the correlation
## make title
tt2 <- paste0(
  'F = ',
  round(corr_anova$`F value`[1], 2),
  '; P ',
  ifelse(corr_anova$`Pr(>F)`[1] < .0001,
         '< 0.0001',
         paste0('= ', round(corr_anova$`Pr(>F)`[1], 3)))
)

# merge and tidy coefficients and confidence intervals
cor_lm_df <- as.data.frame(confint(cor_lm)) %>%
  set_names(c('lower', 'upper')) %>%
  rownames_to_column('term') %>%
  left_join(broom::tidy(cor_lm)) %>%
  filter(term != '(Intercept)') %>%
  mutate(hallmark = as.numeric(fct_reorder(term, (estimate))))

# make graph
corr_lm_plot <-  cor_lm_df %>%
  ggplot(aes(x = hallmark, y = estimate)) +
  geom_point() +
  geom_linerange(aes(ymin = lower, ymax = upper)) +
  scale_x_continuous(breaks = 1:10, labels = paste0('#', 1:10)) +
  coord_flip() +
  labs(x = 'Cancer Hallmarks',
       y = 'Gene-protein Correlation (Estimate)') +
  ggtitle(tt2)

corr_lm_plot
```

This figure reports the results on the linear model and shows that the average gene protein correlation in all hallmarks were significantly different from the random set.

## Cancer hallmarks enrichment in gene and protein based ranks

The next step is to rank the gene products based on gene and protein levels and calculate the enrichment of the cancer hallmark gene sets in the ranked lists.

```{r genes_ernichment}
# enrichment of hallmarks in breast cancer cells (genes) ----
# term_gene <- melt(hallmarks) %>%
#   as_tibble() %>%
#   dplyr::select(term = L1, gene = value)
# 
# stats <- lapply(seq_len(ncol(genes)), function(i) genes[,i])
# stats <- map(stats, sort, decreasing = TRUE)
# names(stats) <- colnames(genes)
# 
# enrich_hallmark_genes <- map_df(stats, function(x) {
#   enrich <- GSEA(x,
#                  pAdjustMethod = 'fdr',
#                  TERM2GENE = term_gene,
#                  pvalueCutoff = 1,
#                  maxGSSize = Inf)
#   as_tibble(enrich)
# }, .id = 'cell_id')
# 
# write_rds(enrich_hallmark_genes, 'data/enrich_hallmark_genes.rds')

enrich_hallmark_genes <- read_rds('data/enrich_hallmark_genes.rds')
```

```{r correlation_enrichment}
# join and tidy the correlation and enrichment variables
correlation_ernichment <- gene_protein_hallmark %>%
  full_join(enrich_hallmark_genes,
            by = c('hallmark' = 'ID',
                   'Var1' = 'cell_id')) %>%
  dplyr::select(hallmark, Var1, value, NES, p.adjust) %>%
  mutate(name = hallmark,
         hallmark = as.numeric(fct_reorder(hallmark, (value))))
```

```{r number_name_hallmarks}
with(select(correlation_ernichment, hallmark, name) %>% unique() %>% arrange(hallmark),
     paste('\\#', hallmark, ', ', name, collapse = '; ', sep = ''))
```

```{r hallmark_enrichment_plot}
# plot the enrichment of the hallmarks
signif_hallmarks <- group_by(correlation_ernichment, hallmark) %>%
  summarise(n = sum(p.adjust < .2),
            max = max(NES))

hallmark_enrichment_plot <- correlation_ernichment %>%
  ggplot(aes(x = (hallmark), y = NES, group = hallmark)) +
  geom_boxplot() +
  annotate('text', x = 1:10,
           y = signif_hallmarks$max + .1,
           label = paste0(signif_hallmarks$n, '*'),
           size = 3) +
  annotate('text', x = 5, y = 1.75, label = 'FDR < 0.2') +
  labs(x = 'Cancer Hallmarks',
       y = 'Enrichment Score') +
  scale_x_continuous(sec.axis = dup_axis(),
                     breaks = 1:10,
                     labels = paste0('#', 1:10)) +
  scale_y_continuous(sec.axis = dup_axis(), 
                     limits = c(.6, 1.8),
                     breaks = c(.8, 1, 1.2, 1.4, 1.6)) +
  my_theme

hallmark_enrichment_plot
```

The figure shows that some hallmarks are significantly enriched some cell lines.

```{r gene_enrichment_points}
# plot hallmark enrichment as points


gene_enrichment_points <- gene_protein_hallmark %>%
  full_join(enrich_hallmark_genes,
            by = c('hallmark' = 'ID',
                   'Var1' = 'cell_id')) %>%
  dplyr::select(hallmark, Var1, value, NES, p.adjust) %>%
  mutate(hallmark = as.numeric(fct_reorder(hallmark, (value)))) %>%
  ggplot(aes(x = hallmark, y = NES, color = p.adjust < .2)) +
  geom_point() +
  scale_x_continuous(breaks = 1:10, labels = paste0('#', 1:10)) +
  labs(x = 'Cancer Hallmarks',
       y = 'Enrichment Score',
       color = 'FDR < 0.2: ') +
  theme(legend.position = 'top',
        legend.key.size = unit(3, "mm"),
        legend.margin =margin(r=0,l=0,t=0,b=0))

gene_enrichment_points
```

This figure shows the same enrichment results in more details.

```{r hallmark_correlation_plot}
# plot the correlation by hallmark
hallmark_groups_test <- map_df(hallmark_groups, function(x) {
  tidy(
    wilcox.test(x,
                hallmark_groups$random,
                alternative = 'greater')
  )
}, .id = 'hallmark') %>%
  filter(hallmark != 'random')

min(hallmark_groups_test$statistic)
max(hallmark_groups_test$p.value)

max_corr <- group_by(correlation_ernichment, hallmark) %>% 
  summarise(max = max(value)) %>%
  pull(max)

hallmark_correlation_plot <- correlation_ernichment %>%
  ggplot(aes(x = hallmark, y = value, group = hallmark)) +
  geom_boxplot() +
  geom_hline(yintercept = mean(hallmark_groups$random), lty = 2, color = 'red') +
  labs(x = 'Cancer Hallmarks', y = 'Gene-protein Correlation') +
  scale_x_continuous(sec.axis = dup_axis(),
                     breaks = 1:10, labels = paste0('#', 1:10)) +
  annotate('text', x = 1, y = .63, size = 3, hjust = 0,
           label = paste('* W >', min(hallmark_groups_test$statistic))) +
  annotate('text', x = 1, y = .58, size = 3, hjust = 0,
           label = '& P < 0.0001') +
  annotate('text', x = 1:10, y = max_corr + .03, size = 3, label = '*') +
  annotate('text', x = 7.5, y = .2, size = 3, color = 'red',
           label = 'Random set average') +
  scale_y_continuous(sec.axis = dup_axis()) +
  my_theme

hallmark_correlation_plot
```

This figure shows that the same hallmarks have a higher gene protein correlations than average, but not all hallmarks follow the enrichment trend.

We then apply the same calculation to the protein-ranked list.

```{r protein_enrichment}
# enrichment of hallmarks in breast cancer cells (proteins)
# stats <- lapply(seq_len(ncol(proteins)), function(i) proteins[,i])
# stats <- map(stats, sort, decreasing = TRUE)
# names(stats) <- colnames(proteins)
# 
# enrich_hallmark_proteins <- map_df(stats, function(x) {
#   enrich <- GSEA(x,
#                  pAdjustMethod = 'fdr',
#                  TERM2GENE = term_gene,
#                  pvalueCutoff = 1,
#                  maxGSSize = Inf)
#   as_tibble(enrich)
# }, .id = 'cell_id')
# 
# write_rds(enrich_hallmark_proteins, 'data/enrich_hallmark_proteins.rds')

enrich_hallmark_proteins <- read_rds('data/enrich_hallmark_proteins.rds')
```

```{r protein_enrichment_points}
# plot hallmark enrichment as points
protein_enrichment_points <- gene_protein_hallmark %>%
  full_join(enrich_hallmark_proteins,
            by = c('hallmark' = 'ID',
                   'Var1' = 'cell_id')) %>%
  dplyr::select(hallmark, Var1, value, NES, p.adjust) %>%
  mutate(hallmark = as.numeric(fct_reorder(hallmark, (value)))) %>%
  ggplot(aes(x = hallmark, y = NES, color = p.adjust < .2)) +
  geom_point() +
  scale_x_continuous(breaks = 1:10, labels = paste0('#', 1:10)) +
  labs(x = 'Cancer Hallmarks',
       y = 'Enrichment Score',
       color = 'FDR < 0.2: ') +
  theme(legend.position = 'top',
        legend.key.size = unit(3, "mm"),
        legend.margin =margin(r=0,l=0,t=0,b=0))

protein_enrichment_points
```

This figure shows the same enrichment results in more details.

```{r enrichment_summary}
# summarize the enrichment scores
summary(enrich_hallmark_genes$NES)
summary(enrich_hallmark_proteins)

# numbers of significantly enriched terms
table(enrich_hallmark_genes$p.adjust < .2)
table(enrich_hallmark_proteins$p.adjust < .2)

# numbers of enriched terms in each cell lines
table(enrich_hallmark_genes$ID, enrich_hallmark_genes$p.adjust < .2)
table(enrich_hallmark_genes$cell_id, enrich_hallmark_genes$p.adjust < .2)
```

## Difference in the hallmarks enrichment between genes and proteins

Here we wanted to know if there is any pattern in the difference between the enrichment based on the gene or the protein ranks.

For every hallmark in each cell line we subtracted the enrichment score which was based on the gene ranking from the one based on the protein ranking.

```{r enrichment_difference}
# calculate the difference between gene and protein enrichment
enrichment_difference <- full_join(enrich_hallmark_genes,
                                   enrich_hallmark_proteins,
                                   by = c('cell_id', 'ID', 'Description')) %>%
  mutate(ID = fct_reorder(ID, NES.x),
         difference = NES.y - NES.x) 
```

```{r enrichment_difference_plot}
# plot the difference in enrichment by hallmark
n <- 10000
set.seed(123)
difference_groups_test <- enrichment_difference %>%
  group_by(ID) %>%
  summarise(p = {
    original <- mean(NES.y - NES.x)
    y <- sample(enrichment_difference$NES.y, 32)
    x <- sample(enrichment_difference$NES.x, 32)
    rand <- replicate(n, mean(y - x))
    mean(abs(rand) >= abs(original))
  },
  min = min(difference))

enrichment_difference_plot <- enrichment_difference %>%
  left_join(difference_groups_test) %>%
  mutate(p = ifelse(p < .0001, '*', '')) %>%
  ggplot(aes(x = difference, y = as.numeric(ID), group = ID)) +
  geom_point() +
  geom_vline(xintercept = 0, lty = 2, color = 'red') +
  geom_text(aes(x = min - .05, label = p)) +
  annotate('text', x = -.6, y = 1.5, label = 'P < 0.0001', hjust = 0) +
  labs(x = 'Difference in enrichment', y = 'Cancer Hallmarks') +
  scale_y_continuous(sec.axis = dup_axis(),
                     breaks = 1:10,
                     labels = paste0('#', 1:10)) +
  scale_x_continuous(sec.axis = dup_axis(),
                     limits = c(-.6, .2),
                     breaks = c(-.6, -.4, -.2, 0, .2)) +
  my_theme

enrichment_difference_plot
```

The figure shows that the difference in enrichment is roughly consistent between the cell lines. That is, for each hallmark, the difference in enrichment scores was higher or lower across the different cell line.

Next, we asked whether these differences in enrichment associate with the division rate of the cell lines under no perturbations?

```{r difference_divison}
# join enrichment difference with division data and summarize
difference_division  <- enrichment_difference %>%
  inner_join(growth, by = c('cell_id' = 'Cell.Name')) %>%
  group_by(cell_id, ID) %>%
  summarise(difference = unique(difference),
            ndr = mean(Nominal.Division.Rate))
```

```{r difference_division_plot}
# plot difference in enrichment against the division rate
difference_division_plot <- difference_division %>%
  ggplot(aes(x = difference, y = ndr)) +
  geom_density2d_filled()+
  labs(x = 'Difference in enrichment',
       y = 'Division Rate') +
  theme(legend.position = 'none') +
  scale_x_continuous(sec.axis = dup_axis(),
                     limits = c(-.45, .25),
                     expand = c(0, 0)) +
  scale_y_continuous(sec.axis = dup_axis(),
                     limits = c(-.1, 2.1),
                     expand = c(0, 0)) +
  geom_vline(xintercept = 0, lty = 2, color = 'white') +
  geom_hline(yintercept = 1, lty = 2, color = 'white') +
  my_theme

difference_division_plot
```

The figure shows that the majority of cells cluster around little difference in enrichment and a moderate division rate. Small groups of cells showed a correlation with the division rate, which we further explored.

```{r difference_lme}
# model the relation between division and difference in enrichment
difference_division_full <- inner_join(enrichment_difference,
                                       growth,
                                       by = c('cell_id' = 'Cell.Name')) %>%
  dplyr::select(cell_id, ID, Nominal.Division.Rate, difference) %>%
  na.omit()

division_lm <- lme(Nominal.Division.Rate ~ difference ,
                   random = ~1|ID,
                   data = difference_division_full)

summary(division_lm)
```

The fixed effect of the difference in enrichment was negative around `r division_lm$coefficients$fixed[2]`

To identify the contribution of the individual hallmarks, we used a separate linear model for each as a group.

```{r difference_lms}
# model by group
division_lm_list <- lmList(Nominal.Division.Rate ~ difference | ID,
                           data = difference_division_full)

division_conf <- map_df(confint(division_lm_list),
                        ~as.data.frame(.x)[2,], .id = 'hallmark')

summary(division_lm_list) 
```

The estimate of the effect of the difference in enrichment on the division rate was consistantly negative and significant.

```{r division_test_plot}
# plot the lm results of division vs difference in enrichment
## make title
tt3 <- paste0(
  'FE = ',
  round(unclass(summary(division_lm))$tTable[2,'Value'], 2),
  '; P ',
  ifelse(corr_anova$`Pr(>F)`[1] < .0001,
         '< 0.0001',
         paste0('= ', round(unclass(summary(division_lm))$tTable[2,'p-value'], 3)))
)

## make figure
division_test_plot <- as.data.frame(unclass(summary(division_lm_list))$coefficients) %>%
  rownames_to_column('hallmark') %>%
  left_join(division_conf) %>%
  mutate(hallmark = as.numeric(fct_reorder(hallmark, (Estimate.difference)))) %>%
  ggplot(aes(x = hallmark, y = Estimate.difference)) +
  geom_point() +
  geom_linerange(aes(ymin = `2.5 %`, ymax = `97.5 %`)) +
  scale_x_continuous(breaks = 1:10, labels = paste0('#', 1:10)) +
  coord_flip() +
  labs(x = 'Cancer Hallmarks',
       y = 'Division Rate (Estimate)') +
  ggtitle(tt3)

division_test_plot
```

The figure shows these estimates and the fixed effect size.

Similarly, we asked whether these difference associate with the growth inhbition in the case of drug treatments.

```{r difference_inhibition}
# join difference in enrichment and growth inhibition
difference_inhibition <- enrichment_difference %>%
  inner_join(growth, by = c('cell_id' = 'Cell.Name')) %>%
  group_by(cell_id, ID, Small.Molecule.Name) %>%
  summarise(difference = unique(difference),
            grmax = mean(Normalized.Growth.Rate.Inhibition.Value))
```

```{r difference_inhibition_plot}
# plot difference in enrichment against the growth inhibition
difference_inhibition_plot <- difference_inhibition %>%
  ggplot(aes(x = difference, y = grmax)) +
  geom_density2d_filled() +
  geom_vline(xintercept = 0, lty = 2, color = 'white') +
  geom_hline(yintercept = 0, lty = 2, color = 'white') +
  geom_hline(yintercept = 1, lty = 2, color = 'white') +
  labs(x = 'Difference in enrichment',
       y = 'Growth Inhibition') +
  theme(legend.position = 'none') +
  scale_x_continuous(sec.axis = dup_axis(),
                     limits = c(-.47, .25),
                     expand = c(0, 0)) +
  scale_y_continuous(sec.axis = dup_axis(),
                     limits = c(-.9, 1.49),
                     expand = c(0, 0)) +
  my_theme

difference_inhibition_plot
```

No clear correlation between the two variables could be discerned from the figure. Differences in enrichment at all levels were associated with moderate inhibition of growth as a result of drug perturbations.

To test these association in a different way we do the following
```{r test_difference_with_growth}
# difference in enrichment and the division rate
cor(difference_division$difference, difference_division$ndr)
cor.test(difference_division$difference, difference_division$ndr)

# difference in enrichment and growth inhibition with perturbations
cor(difference_inhibition$difference, difference_inhibition$grmax)
cor.test(difference_inhibition$difference, difference_inhibition$grmax)
```

To put both to gather we join the data in one big data.frame and aggregate the effects.

```{r difference_growth}
# join and tidy difference in enrichment and growth data
difference_growth <- enrichment_difference %>%
  inner_join(growth, by = c('cell_id' = 'Cell.Name')) %>%
  group_by(cell_id, ID, Small.Molecule.Name) %>%
  summarise(difference = unique(difference),
            grmax = min(Normalized.Growth.Rate.Inhibition.Value),
            ndr = mean(Nominal.Division.Rate)) %>%
  ungroup() %>%
  group_by(ID) %>%
  summarise('Division_Rate_cor' = cor(difference, ndr),
            'Division_Rate_p' = cor.test(difference, ndr)$p.value,
            'Growth_Inhibition_cor' = cor(difference, grmax),
            'Growth_Inhibition_p' = cor.test(difference, grmax)$p.value) %>%
  mutate(ID = as.numeric(fct_reorder(ID, `Division_Rate_cor`)),
         ID = paste0('#', as.factor(as.numeric(ID))),
         ID = factor(ID, levels = paste0('#', 1:10))) %>%
  gather(type1, cor, Division_Rate_cor, Growth_Inhibition_cor) %>%
  gather(type2, pvalue, Division_Rate_p, Growth_Inhibition_p) %>%
  mutate(type1 = str_replace_all(type1, '_|cor', ' ')) %>%
  mutate(type2 = str_replace_all(type2, '_|p', ' ')) %>%
  filter(type1 == type2)
```

```{r division_growth}
# plot the correlation of difference to growth
division_growth <- difference_growth %>%
  mutate(p = ifelse(pvalue < .0001, '*', '')) %>%
  ggplot(aes(x = cor, y = ID)) +
  geom_col() +
  geom_text(aes(x = cor - .05, label = p), vjust = 1) +
  facet_wrap(~type1) +
  annotate('text', x = -.55, y = 10, label = 'P < 0.0001', hjust = 0, size = 3) +
  scale_x_continuous(breaks = -c(0, .25, .5), labels = c('0', '-0.25', '-0.5')) +
  labs(x = "Pearson's Correlation",
       y = 'Cancer Hallmarks') +
  my_theme +
  theme(panel.spacing = unit(-.2, 'mm'),
        strip.background = element_blank())

division_growth
```

The figure shows that strong correlations exist between the difference in enrichment and the division rates, but not the growth inhibition.

## range of enrichment of the hallmarks

Now, how about the range of enrichment based on either genes or proteins ranking? Do these ranges, max - min, correlate with the growth variables?

To find out, we calculated the range by subtracting the minimum from the maximum enrichment score of the hallmarks which were based on either the genes or the proteins ranking.

```{r enrichment_range}
# calculate the range of enrichment
enrichment_range <- enrichment_difference %>%
  inner_join(growth, by = c('cell_id' = 'Cell.Name')) %>%
  ungroup() %>%
  group_by(cell_id, Small.Molecule.Name) %>%
  summarise(
    gene_min = min(NES.x),
    gene_max = max(NES.x),
    gene_diff = gene_max - gene_min,
    protein_min = min(NES.y),
    protein_max = max(NES.y),
    protein_diff = protein_max - protein_min,
    ndr_max = max(Relative.Cell.Count),
    ndr_med = median(Nominal.Division.Rate),
    grmax_med = median(Normalized.Growth.Rate.Inhibition.Value)
  ) %>%
  unique()
```

And then calculated the correlation with the medium growth variables for each cell line.

```{r range_ndr_correaltion}
# range of enrichment correlation with grmax_med
range_ndr_correaltion <- enrichment_range %>%
  group_by(Small.Molecule.Name) %>%
  summarise(Genes = cor(gene_diff, ndr_med),
            Proteins = cor(protein_diff, ndr_med))
```

```{r enrichment_range_plot}
# plot the range of enrichment
range_ndr_ttest <- t.test(range_ndr_correaltion$Genes,
                          range_ndr_correaltion$Proteins,
                          alternative = 'greater')
range_ndr_ttest

enrichment_range_plot <- range_ndr_correaltion %>%
  gather(key, value, -Small.Molecule.Name) %>%
  ggplot(aes(x = value, fill = key)) +
  geom_bar(stat="bin", aes(y=..density..), alpha = .5) +
  geom_vline(xintercept = range_ndr_ttest$estimate,
             color = c('red', 'blue'),
             lty = 2) +
  geom_segment(aes(x = range_ndr_ttest$estimate[1],
                   xend = range_ndr_ttest$estimate[2],
                   y = 2, yend = 2),
               arrow = arrow(length = unit(2, 'mm'))) +
  geom_segment(aes(xend = range_ndr_ttest$estimate[1],
                   x = range_ndr_ttest$estimate[2],
                   y = 2, yend = 2),
               arrow = arrow(length = unit(2, 'mm'))) +
  annotate('text', x = -.45, y = 3, size = 3, hjust = 0,
           label = paste('t =', round(range_ndr_ttest$statistic, 1))) +
  annotate('text', x = -.45, y = 2.5, size = 3, hjust = 0,
           label = paste('P < 0.0001')) +
  labs(x = "Pearson's Correlation",
       y = 'Percent of drugs',
       fill = '') +
  theme(legend.position = c(.5,.9),
        legend.background = element_blank()) +
  scale_x_continuous(sec.axis = dup_axis(),
                     breaks = c(-.6,-.3, 0, .3)) +
  scale_y_continuous(sec.axis = dup_axis()) +
  my_theme

enrichment_range_plot
```

The figure shows that the range of enrichment based on the proteins had strong and negative correlation with the growth inhibition under drug perturbations. That is, cells that had a wide range of enrichment of proteins were less likely to be inhibited by drug treatments.

We then tested the difference between the cumulative distributions of the enrichment ranges of gene based and protein based ranking.

```{r test_range_ndr}
set.seed(123)
ks.test(range_ndr_correaltion$Genes, 'rnorm', alternative = 'less')
ks.test(range_ndr_correaltion$Proteins, 'rnorm', alternative = 'greater')

range_ndr_test <- ks.test(range_ndr_correaltion$Genes, 
                          range_ndr_correaltion$Proteins)

range_ndr_test
```

```{r range_ndr_test_plot}
cdf1 <- ecdf(range_ndr_correaltion$Genes) 
cdf2 <- ecdf(range_ndr_correaltion$Proteins)

minMax <- seq(min(range_ndr_correaltion$Genes, range_ndr_correaltion$Proteins, na.rm = TRUE),
              max(range_ndr_correaltion$Genes, range_ndr_correaltion$Proteins, na.rm = TRUE),
              length.out=length(range_ndr_correaltion$Genes)) 

x0 <- minMax[which( abs(cdf1(minMax) - cdf2(minMax)) == max(abs(cdf1(minMax) - cdf2(minMax))) )] 
y0 <- cdf1(x0) 
y1 <- cdf2(x0) 

tt4 <- paste0(
  'D^+ = ',
  round(range_ndr_test$statistic, 2),
  '; P ',
  ifelse(range_ndr_test$p.value < .0001,
         '< 0.0001',
         paste0('= ', round(range_ndr_test$p.value, 3)))
)

range_ndr_test_plot <- range_ndr_correaltion %>%
  gather(key, value, -Small.Molecule.Name) %>%
  ggplot() +
  stat_ecdf(aes(x = value, color = key)) +
  geom_segment(aes(x = x0[1], y = y0[1], xend = x0[1], yend = y1[1]),
               linetype = "dashed" ) +
  geom_point(aes(x = x0[1] , y= y0[1]), size=3) +
  geom_point(aes(x = x0[1] , y= y1[1]), size=3) +
  ggtitle(tt4) + 
  labs(x = 'Range-division correlation',
       y = 'ECDF',
       color = '') +
  theme(legend.position = c(.47,.5),
        legend.background = element_blank())

range_ndr_test_plot
```

This figure shows the big difference between the cumulative distribution of both curves.

# Final figures

```{r final_figures}
# hallmark_enrichment.png
plot_grid(hallmark_enrichment_plot,
          enrichment_difference_plot,
          nrow = 1,
          scale = .9,
          labels = 'AUTO',
          label_fontface = 'plain') %>%
  ggsave(plot = .,
         filename = 'output/manuscript/figures/hallmark_enrichment.png',
         width = 6, height = 3)

# enrichment_test.png
plot_grid(gene_enrichment_points,
          protein_enrichment_points,
          nrow = 1,
          scale = .9,
          labels = 'AUTO',
          label_fontface = 'plain') %>%
  ggsave(plot = .,
         filename = 'output/manuscript/figures/enrichment_test.png',
         width = 6, height = 3)

# gene_protein_correlation.png
plot_grid(gene_protein_corr_plot,
          hallmark_correlation_plot,
          nrow = 1,
          scale = .9,
          labels = 'AUTO',
          label_fontface = 'plain') %>%
  ggsave(plot = .,
         filename = 'output/manuscript/figures/gene_protein_correlation.png',
         width = 6, height = 3)

# gene_protein_correlation_tests.png
plot_grid(gene_protein_test_plot,
          corr_lm_plot,
          nrow = 1,
          scale = .9,
          labels = 'AUTO',
          label_fontface = 'plain') %>%
  ggsave(plot = .,
         filename = 'output/manuscript/figures/gene_protein_correlation_tests.png',
         width = 6, height = 3)

# division_inhibition.png
plot_grid(difference_division_plot,
          difference_inhibition_plot,
          nrow = 1,
          scale = .9,
          labels = 'AUTO',
          label_fontface = 'plain') %>%
  ggsave(plot = .,
         filename = 'output/manuscript/figures/division_inhibition.png',
         width = 6, height = 3)

# difference_range.png
plot_grid(division_growth,
          enrichment_range_plot,
          nrow = 1,
          scale = .9,
          labels = 'AUTO',
          label_fontface = 'plain') %>%
  ggsave(plot = .,
         filename = 'output/manuscript/figures/difference_range.png',
         width = 6, height = 3)

# difference_range_tests.png
plot_grid(division_test_plot,
          range_ndr_test_plot,
          nrow = 1,
          scale = .9,
          labels = 'AUTO',
          label_fontface = 'plain') %>%
  ggsave(plot = .,
         filename = 'output/manuscript/figures/difference_range_tests.png',
         width = 6, height = 3)
```

